An Analysis of the Use of Average Centipawn Loss to Detect Cheating in Chess

### 2022-11-15 Manuscript/manuscript.tex
fixed citations

The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. Cheating in online chess is very easy. One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next. Cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.

@@

Item 2 requires an assessment of move quality, which is usually done in centipawns. The centipawn is a unit of measure used in chess to quantify advantage. A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions. Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine. Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.

@@

The data used was provided by Lichess.com, a free and open-source internet chess server that holds the records of many professional games. It is a comprehensive list of every professional game Hans Niemann has played, both virtually and OTB, from March of 2018 to September of 2022 and is provided in both .csv and .pkl format. Some of the relevant information it contains are what color Niemann and his opponent played and what their respective ELO scores and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.). The main information of concern with regard to the entirety of the data is Niemann's ELO score and calculated ACL score of each game of each particular point in time, as this is what will be used to plot. All other games that included information where Niemann was not playing were dropped from the original dataset provided by Lichess.

@@

% how to make individual images that are different resolutions the same size?
\begin{figure}
    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.2]{niemann_rating.png}
    \caption{Hans Niemann's ELO ratings over time.}
    \label{fig:sfig1}
    \end{subfigure}

    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.45]{carlsen_rating.png}
    \caption{Magnus Carlsen's ELO ratings over time.}
    \label{fig:sfig2}
    \end{subfigure}
\label{fig:fig}
\end{figure}

>>>>

The first game was played “over the board” (OTB) in contrast to “virtual chess” which is hosted online. Cheating in online chess is very easy. One simply needs to run a chess engine (like Stockfish) and execute the moves from their game into the engine as they played, following up with whichever move it suggested they play next \citep{kenregan2010}. Cheating OTB is not as simple. Players are commonly required to pass through a metal detector and are not permitted to bring phones or any other electronic devices with them.

@@

Item 2 requires an assessment of move quality, which is usually done in centipawns. The centipawn is a unit of measure used in chess to quantify advantage. A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions . Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine. Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.

@@

The data used was provided by Lichess.com (via kaggle) \citep{zaidqureshi2022}, a free and open-source internet chess server that holds the records of many professional games. It is a comprehensive list of every professional game Hans Niemann has played, both virtually and OTB, from March of 2018 to September of 2022 and is provided in both .csv and .pkl format. Some of the relevant information it contains are what color Niemann and his opponent played and what their respective ELO scores and calculated ACL scores were along with some other unimportant details for this analysis (e.g., ACL score of Niemann's opponent, tournament name and event round, etc.). The main information of concern with regard to the entirety of the data is Niemann's ELO score and calculated ACL score of each game of each particular point in time, as this is what will be used to plot. All other games that included information where Niemann was not playing were dropped from the original dataset provided by Lichess.

@@

% how to make individual images that are different resolutions the same size?
\begin{figure}
    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.2]{niemann_rating.png}
    \caption{Hans Niemann's ELO ratings over time. \citep{hansniemannrating}}
    \label{fig:sfig1}
    \end{subfigure}

    \begin{subfigure}{.5\textwidth}
    \includegraphics[scale=0.45]{carlsen_rating.png}
    \caption{Magnus Carlsen's ELO ratings over time. \citep{hansniemannrating}}
    \label{fig:sfig2}
    \end{subfigure}
\label{fig:fig}
\end{figure}
